Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2013-01-22 HRUIonGPU/jrnl_article.tex
finally started

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

@@

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. We can therefore easily remove beams where no interference is present by removing rows from $\mat{B}$.

@@

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray for the $k^{th}$ beam is:
\begin{align}\label{eq:bs_formula}
x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
\end{align}
Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
\label{fig:benchBS5}}}
\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector. Execution time is plotted as a function of subarray length $L$. (a) Shows the benchmark for a 64 elements array. (b) Shows the benchmark for a 96 elements array.}
\label{fig:benchBS}
\end{figure*}

@@

For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers. The cardiac recording only has $30\%$ lateral over sampling.

\section{Trading resolution for speed}
Add a comparison between beamspace and full capon using small L. Plot system resolution vs $L$ and $N_b$ for the two methods. ???
Fig \ref{fig:speed_res_trade}.

\begin{figure}
\centering
\includegraphics[width=3in]{gfx/speed_res_trade.png}
\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace.}
\label{fig:speed_res_trade}
\end{figure}

\section{Discussion}\label{sec:dis}
From Fig. \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 100 ms (10 fps). Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al. \cite{Chen2011}. For the 32 element array in Fig. (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$). Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32. The frame rate for a 64 element array is then increased from 10 to 44 fps. Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.

A major part of the total execution time comes from solving systems of linear equations. Further research should therefore focus on making this contribution smaller. This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}. %When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
As for now, we have shown that the Capon beamformer can be implemented in a GPU framework. And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig. (\ref{fig:bench}) are real time. If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively. There is some distance to cover before that goal is reached for all values of $L \le M/2$. However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$. Another option to reach real-time execution is to use several GPUs simultaneously. Note that multi-line acquisition will increase the frame rate requirement accordingly.

>>>>

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%

%\documentclass[journal]{IEEEtran}

% Single column review mode:
\documentclass[12pt,journal,onecolumn]{IEEEtran}
\newcommand\CLASSINPUTbaselinestretch{2}
\RequirePackage{calc}
\RequirePackage{fp} 
% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
% Adding the 2cm gutter width to the odd/even side margins:
\setlength\hoffset{0pt}
\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side

%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

@@

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. We can therefore remove beams where no interference is present by removing rows from $\mat{B}$.

@@

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray for the $k^{th}$ beam is:
\begin{align}\label{eq:bs_formula}
x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
\end{align}
Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We denote this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
\label{fig:benchBS5}}}
\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector. Execution time is plotted as a function of subarray length $L$. (a) Shows the benchmark for a 64 elements array. (b) Shows the benchmark for a 96 elements array. (TODO: Change this to be a function of $N_b$)}
\label{fig:benchBS}
\end{figure*}

@@

For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers as they move. The cardiac recording only has $30\%$ lateral over sampling.

\section{Trading resolution for speed}
As we see from Fig. \ref{fig:bench} and \ref{fig:benchBS}, the execution time is reduced when either $L$ or $N_b$ is reduced. The question is how this impact the image quality. In Fig. \ref{fig:speed_res_trade} the minimum distance between the two point targets in Fig. \ref{fig:phantom} (while preserving a 6dB lower level in between) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon). For high values the two methods are equal, but for lower values BS Capon provides higher resolution than what is possible with small subarrays. For this result, $d=0.01$ has been used on both methods.

\begin{figure}
\centering
\includegraphics[width=3in]{gfx/speed_res_trade.png}
\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension (based on Fig. \ref{fig:phantom}). As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig. \ref{fig:bench} and \ref{fig:benchBS}). When $N_b$ is varied, $L$ is fixed on 32.}
\label{fig:speed_res_trade}
\end{figure}

\section{Discussion}\label{sec:dis}
We have presented a parallel implementation of both the Capon og BS Capon beamformer.

From Fig. \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 100 ms (10 fps). Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al. \cite{Chen2011}. For the 32 element array in Fig. (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$). Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32. The frame rate for a 64 element array is then increased from 10 to 44 fps. Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.

A major part of the total execution time comes from solving systems of linear equations. This limitation is targeted by reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.

We have shown that the Capon beamformer can be implemented in a GPU framework. And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig. (\ref{fig:bench}) are real time. If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively. There is some distance to cover before that goal is reached for all values of $L \le M/2$. However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$. Another option to reach real-time execution is to use several GPUs simultaneously. Note that multi-line acquisition will increase the frame rate requirement accordingly.
